---
title: "Week_7_MTOF_data"
author: "Gheeda Mourtada"
date: "17/06/2021"
output: pdf_document
---

# Outline

- Points Drawn from _A Structured Tumor-Immune Microenvironment in Triple Negative Breast Cancer Revealed by Multiplexed Ion Beam Imaging_

- Creating `SpatialExperiment` for MIBI-TOF data

- Using `SpatialExperiment` object for creating heatmaps


## Introductory Notes

_MIBI-TOF:_ multiplexed ion beam imaging by time-of-flight 

**Goal:** understanding tumor-immune environment; learning more about immune phenotypes, compositions, and interactions with tumor cells.

**Results:** quantified _in-situ_ expression of 36 proteins in 41 patients.


\newpage

# SpatialExperiment for MIBI-TOF data

```{r include=FALSE}
library(dplyr)
library(plyr)
library(raster)
library(rgdal)
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(SpatialExperiment)
```
```{r}
load("~/Downloads/mibiSCE.rda")
mibi<-mibi.sce
```

\newpage

# Creating `SpatialExperiment` object from `SingleCellExperiment` object

```{r}
SE <- SpatialExperiment(
  rowData = rowData(mibi),
  colData = colData(mibi),
  assays = assay(mibi))
SE
```

\newpage

# Getting spatial information fron tiff files

\footnotesize
```{r}
setwd("~/Downloads")
tiff_file_list<- list.files(path="TNBC_shareCellData/",
                            pattern = ".tiff")[1:41]
tiff_file_ID <- as.list(seq_along(tiff_file_list))

SpatialCoordGetter <- function(id){
  
  # Similarly to the previously created "for" loop, we extract the sample IDs from the tiff files and create raster object from it
  sample_id <- as.numeric(gsub("p", "", gsub("_labeledcellData.tiff", "", tiff_file_list[id])))
  str_name <- paste("TNBC_shareCellData/",tiff_file_list[id],
                    sep = "")
  r <- raster(str_name)
  
  # creating dataframe which includes spatial information
  df_rP <- data.frame(rasterToPoints(r))
  
  # changing "value" column to "cellLabel"
  colnames(df_rP)[3] <- "cellLabel"
  
  # compute centroid x coordinate of each cell; first column of our dataframe (median of each cell using "cellLabel" as the list of grouping elements)
  centroid_x <- aggregate(df_rP[,1], 
    by = list(df_rP[,3]), FUN = median) %>%
     dplyr::rename('cellLabel' = Group.1)  
  # new name = old name; we get a data frame with cell label as the first column and x centroid as the second

  # compute centroid y coordinate of each cell; second column of our dataframe (similarly to calculating x centroid)
  centroid_y <- aggregate(df_rP[,2], 
    by = list(df_rP[,3]), FUN = median) %>% 
    dplyr::rename('y' = x, 'cellLabel' = Group.1) #renaming columns of dataframe

  # creating a data frame with cellLabel, x, and y, and adding a sample id column
  centroid_xy <- left_join(centroid_x, centroid_y, 
                           by = "cellLabel")
  
  centroid_xy <- centroid_xy %>% mutate(sample_id = rep(sample_id,
                                                        nrow(centroid_xy)))
  
  return(centroid_xy)
}

#binding the data frames of centorid of cell labels for all subjects
centroid_xy_df <- map_dfr(tiff_file_ID, SpatialCoordGetter)

# adding centroid_x and centroid_y to SE (spatial experiment object) while preserving the order of SampleID and cellLabelInImage
mibi_df <- data.frame(colData(SE))
mibi_df <- left_join(mibi_df,
                     centroid_xy_df,
                     by = c("SampleID" = "sample_id", "cellLabelInImage" = "cellLabel"))

SE$centroid_x <- mibi_df$x
SE$centroid_y <- mibi_df$y

```
\normalsize

\newpage

# Adding `spatialCoords` to our previously created `SpatialExperiment` object

```{r}
spatial_coords <- as.matrix(data.frame(
  x = SE$centroid_x, 
  y = SE$centroid_y
  ))

spatialCoords(SE) <- spatial_coords
SE
save(SE,file = "se.rds")
```

\newpage

# Exploring the `SpatialExperiment` object

```{r}
rowMeans(assay(SE))
```

\newpage

```{r}
rowSds(assay(SE))
```

The channels that were collected across cells in the experiment

```{r}
rownames(SE)
```

\newpage

Using the binary attribute is_protein for data filtering

```{r}
SE_proteins <- SE[rowData(SE)$is_protein == 1,]
SE_proteins
```

Removing the `SE` object from environment and using SE_proteins` moving forward

```{r}
rm(SE)
```

Identifying abundance of each cell type in `tumor_group`

```{r}
round(table(SE_proteins$tumor_group)/ncol(SE_proteins),2)
```

Identifying abundance of each cell type in `immune_group`

```{r}
round(table(SE_proteins$immune_group)/ncol(SE_proteins),2)
```

\newpage

Keratin +ve tumor cells and immune cells

```{r}
table(colData(SE_proteins)$tumor_group)
```

```{r}
table(colData(SE_proteins)$tumor_group, colData(SE_proteins)$immune_group)
```

\footnotesize
```{r}
SE_proteins <- SE_proteins[,
                                       colData(SE_proteins)$tumor_group %in% c("Immune",
                                                                                       "Keratin-positive tumor")]
```
\normalsize

\newpage

# Processing data using `SpatialExperiment` object

```{r}
SE <- SE_proteins
SE
```


Dropping columns with no protein expression

```{r}
SE <- SE[, colSums(assay(SE)) > 0]
SE
```

\newpage

Dropping rows with no expression across columns

```{r}
SE <- SE[rowSums(assay(SE)) > 0, ]
SE
```


```{r}
summary(as.vector(assay(SE)))
hist(as.vector(assay(SE)), breaks = 100)
```

\newpage

Removing outlier expression

```{r}
rm_outliers <- apply(assay(SE), 2, function(x){ sum(abs(x) > 5) >= 1 })
SE <- SE[, which(!rm_outliers)]
SE
```

Adding cell IDs

```{r}
colData(SE)$cell_id <- paste0("mibi_", seq(1, dim(SE)[2]))
colnames(SE) <- colData(SE)$cell_id
colData(SE)$DONOR_NO <- colData(SE)$DONOR_NO %>% as.character()
SE
```

\newpage

# Heatmap from our `SpatialExperiment` Object

## Plotting two categorical variables

```{r eval=FALSE}
col_SE<-data.frame(colData(SE))

anno_col = data.frame( TumorGroup=col_SE$tumor_group,
                       ImmuneGroup=col_SE$immune_group,
                       DonorNO=col_SE$DONOR_NO)

rownames(anno_col)<-colnames(assay(SE))

pheatmap::pheatmap(assay(SE),
                   annotation_col = anno_col,
                   cluster_rows = FALSE, 
                   cluster_cols = FALSE, 
                   fontsize_row = 8,
                   show_colnames = FALSE)
```

\newpage

```{r echo=FALSE}
col_SE<-data.frame(colData(SE))

anno_col = data.frame( TumorGroup=col_SE$tumor_group,
                       ImmuneGroup=col_SE$immune_group,
                       DonorNO=col_SE$DONOR_NO)

rownames(anno_col)<-colnames(assay(SE))

p<-pheatmap::pheatmap(assay(SE),
                   annotation_col = anno_col,
                   cluster_rows = FALSE, 
                   cluster_cols = FALSE, 
                   fontsize_row = 8,
                   show_colnames = FALSE)
p
ggsave("SE.png",p)
```

\newpage

## Plotting spatial information for each sample from the `SpatialExperiment` object

\footnotesize
```{r}
manual_col_sample_id <- c("B"="#1f77b4",
                          "CD3"="#aec7e8",
                          "CD4"="#ff7f0e",
                          "CD8"="#ffbb78",
                          "DC"="#2ca02c",
                          "DC/Mono"="#98df8a",
                          "Macrophages"="#c49c94",
                          "Mono/Neu"="#ff9896",
                          "Neutrophils"="#9467bd",
                          "NK"="#c5b0d5",
                          "Other immune"="#8c564b",
                          "Tregs"="#e377c2",
                          "Tumor"="#d62728",
                          "Background"="black")

tiff_file_list<- list.files(path="TNBC_shareCellData/",
                            pattern = ".tiff")
id<-c(1:41)

for (val in id){
str_name_id<- paste("TNBC_shareCellData/",tiff_file_list[val], sep = "")

sample_id_each <- as.numeric(gsub("p", "", gsub("_labeledcellData.tiff", "", tiff_file_list[val])))
r <- raster(str_name_id)
mibi_sample_id <- SE[, SE$SampleID == sample_id_each]

saveRDS(mibi_sample_id, paste(sample_id_each,".rds"))

# Reading one of the saved SingleCellExperiment objects

sce_file_list<- list.files(pattern = ".rds")

df_rP <- data.frame(rasterToPoints(r))
colnames(df_rP) <- c("X", "Y", "values")
head(df_rP)

noise_not_in_mibi <- unique(df_rP$values[!df_rP$values %in% mibi_sample_id$cellLabelInImage])

# Adding cell_type column
cell_type <- ifelse(mibi_sample_id$immune_group != "not immune", mibi_sample_id$immune_group, "Tumor")
mibi_sample_id$cell_type <- cell_type

# Mapping values from those that are in noise_not_in_mibi to Background
cell_label_with_bg <- mapvalues((df_rP$values), 
                                from = noise_not_in_mibi, 
                                to = rep("Background", length(noise_not_in_mibi)))

# Mapping values which are not Background to cell_type
cell_label_with_bg <- mapvalues((cell_label_with_bg), 
                                from = mibi_sample_id$cellLabelInImage, 
                                to = mibi_sample_id$cell_type)

df_rP$cell_label_with_bg <- cell_label_with_bg

df_rP$cell_label_with_bg <- factor(df_rP$cell_label_with_bg)

df_rP$cell_label_with_bg <- factor(df_rP$cell_label_with_bg,
                                   levels = c(levels(df_rP$cell_label_with_bg)[-which(levels(df_rP$cell_label_with_bg) == "Background")], "Background"))

df_rP$subjectID <- sample_id_each 

assign( paste("df_rP_sample_ID", val, sep = "_") , df_rP)  

p<-ggplot(NULL) + 
  geom_raster(data = df_rP, 
              aes(X, Y, fill = cell_label_with_bg)) +
  theme_minimal() +
  scale_fill_manual(values = manual_col_sample_id) +
  coord_quickmap() +
  theme(aspect.ratio = 1, legend.position = "right") +
  labs(fill = "Cell types")

ggsave(paste0("df_rP_sample_ID_plot", val, ".png"), p)

}
```
\normalsize


\newpage

# References

[A Structured Tumor-Immune Microenvironment in Triple Negative Breast Cancer Revealed by Multiplexed Ion Beam Imaging](https://www.cell.com/action/showPdf?pii=S0092-8674%2818%2931100-0)

[Building SpatialExperiment objects](https://bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html)

Creating SpatialExperiment from SingleCellExperiment and images _Dr. Jeganathan_

[Efficiently bind multiple data frames by row and column](https://dplyr.tidyverse.org/reference/bind.html)

[Raster data manipulation](https://rspatial.org/raster/spatial/8-rastermanip.html)

[Raster usage for accessing coordinate labels](https://github.com/BIRSBiointegration/Hackathon/tree/master/sc-targeted-proteomics/spatial_data)

Integrating proteomics data _Dr. Jeganathan_


[MIBI Datasets](https://www.angelolab.com/mibi-data)

[Pheatmap Draws Pretty Heatmaps](https://towardsdatascience.com/pheatmap-draws-pretty-heatmaps-483dab9a3cc)

[SpatialExperiment-methods: Methods for spatial attributes](https://rdrr.io/github/drighelli/VisiumExperiment/man/SpatialExperiment-methods.html)
