---
title: "Spatial Statistical Methods for Spatial Omic Data"
author: "Gheeda Mourtada"
date: "21/07/2021"
output: beamer_presentation
---

# Outline

- Image segmentation
- Spatial point pattern analysis
- UMAP on protein expression

----

# Image Segmentation

## Visualizing _tiff_ images

```{r include=FALSE}
library("SpatialExperiment")
library("EBImage")
library("readr")
library("dplyr")
library("ggplot2")
library("spatstat")
library("geometry")
library("RColorBrewer")
library("raster")
library("genefilter")
```

\footnotesize
```{r echo=FALSE, out.width="70%"}
sub_1_tiff = readImage("/Users/mourtadg/Desktop/NSERC/TNBC_shareCellData/p1_labeledcellData.tiff")
display(sub_1_tiff)
```
\normalsize

-----

# Example from _Modern Statistics for Modern Biology_ Chapter 11

```{r}
library("MSMB")
imagefiles = system.file("images",
                         c("image-DAPI.tif",
  "image-FITC.tif", "image-Cy3.tif"),
  package="MSMB")
cells = readImage(imagefiles)
```

<!-- ----- -->

<!-- ```{r} -->
<!-- display(cells) -->
<!-- ``` -->

-----


```{r}
apply(cells, 3, range)
cells[,,1]   = 32 * cells[,,1]
cells[,,2:3] = 16 * cells[,,2:3]
apply(cells, 3, range)
```

-----

## Noise reduction by smoothing

```{r}
cellsSmooth = Image(dim = dim(cells))
sigma = c(1, 3, 3)
for(i in seq_along(sigma))
  cellsSmooth[,,i] = filter2( cells[,,i],
         filter = makeBrush(size = 51, shape = "gaussian",
                            sigma = sigma[i]) )
```

-----

## Adaptive thresholding

```{r include=FALSE}
disc = makeBrush(21, "disc")
disc = disc / sum(disc)
```

\footnotesize
```{r}
nucThresh =
  (cellsSmooth[,,1] - filter2(cellsSmooth[,,1], disc) > 0.02)
```
\normalsize

-----

# Morphological opening

```{r}
nucOpened = EBImage::opening(nucThresh,
                  kern = makeBrush(5, shape = "disc"))
```

----

# Segmentation into objects

```{r}
nucSeed = bwlabel(nucOpened)
table(nucSeed)
```

-----

# Issues: improper separations and holes

```{r echo=FALSE}
display(colorLabels(nucSeed))
```

```{r include=FALSE}
nucMask = cellsSmooth[,,1] - filter2(cellsSmooth[,,1], disc) > 0
nucMask = fillHull(nucMask)
nuclei = propagate(cellsSmooth[,,1], nucSeed, mask = nucMask)
```

-----

# Voronoi Tessellation

\footnotesize
```{r include=FALSE}
zeros        = Image(dim = dim(nuclei))
voronoiExamp = propagate(seeds = nuclei, x = zeros, lambda = 100)
voronoiPaint = paintObjects(voronoiExamp, 1 - nucOpened)
```
\normalsize

```{r echo=FALSE}
display(voronoiPaint)
```

-----

```{r}
hist(log(cellsSmooth[,,3]) )
```

-----

\footnotesize
```{r}
hist(log(cellsSmooth[,,3]), xlim = -c(3.6, 3.1), breaks = 300)
```
\normalsize

-----

```{r include=FALSE}
bgPars = function(x) {
  x    = log(x)
  loc  = half.range.mode( x )
  left = (x - loc)[ x < loc ]
  wid  = sqrt( mean(left^2) )
  c(loc = loc, wid = wid, thr = loc + 6*wid)
}
cellBg = apply(cellsSmooth, MARGIN = 3, FUN = bgPars)

hist(log(cellsSmooth[,,3]), xlim = -c(3.6, 3.1), breaks = 300)
abline(v = cellBg[c("loc", "thr"), 3], col = c("brown", "red"))

cytoplasmMask = (cellsSmooth[,,2] > exp(cellBg["thr", 2])) |
       nuclei | (cellsSmooth[,,3] > exp(cellBg["thr", 3]))

cellbodies = propagate(x = cellsSmooth[,,3], seeds = nuclei,
                       lambda = 1.0e-2, mask = cytoplasmMask)
cellsColor = rgbImage(red   = cells[,,3],
                      green = cells[,,2],
                      blue  = cells[,,1])

nucSegOnNuc  = paintObjects(nuclei, tgt = toRGB(cells[,,1]),
                            col = "#ffff00")
nucSegOnAll  = paintObjects(nuclei, tgt = cellsColor,
                            col = "#ffff00")
cellSegOnAll = paintObjects(cellbodies, tgt = nucSegOnAll,
                            col = "#ff0080")
```

```{r echo=FALSE}
display(cellbodies)
```

-----

```{r echo=FALSE}
display(cellsColor)
```

-----

```{r echo=FALSE}
display(nucSegOnNuc)
```

-----

```{r echo=FALSE}
display(nucSegOnAll)
```

-----

```{r echo=FALSE}
display(cellSegOnAll)
```

-----

# Spatial point processes

```{r}
load("/Users/mourtadg/Desktop/NSERC/se.rds")
```

```{r include=FALSE}
SE
SE_proteins <- SE[rowData(SE)$is_protein == 1,]
SE_proteins <- SE_proteins[,colData(SE_proteins)$tumor_group %in% c("Immune", "Keratin-positive tumor")]
SE <- SE_proteins
SE <- SE[, colSums(assay(SE)) > 0]
rm_outliers <- apply(assay(SE), 2, function(x){ sum(abs(x) > 5) >= 1 }) 
SE <- SE[, which(!rm_outliers)]
colData(SE)$cell_id <- paste0("mibi_", seq(1, dim(SE)[2]))
colnames(SE) <- colData(SE)$cell_id
colData(SE)$DONOR_NO <- as.character(colData(SE)$DONOR_NO)
```

```{r}
SE
```

-----

## Subsetting a subject

```{r}
subject_1<-SE[,colData(SE)$SampleID==1]
subject_1
```

-----

```{r}
head(spatialCoords(subject_1))
```

-----

```{r}
cell_type <- ifelse(subject_1$immune_group != "not immune",
                    subject_1$immune_group, "Tumor") 
subject_1$cell_type <- cell_type

spatial_coor<-as.data.frame(spatialCoords(subject_1))
spatial_coor$cell_type<-cell_type
head(spatial_coor)
```

-----

```{r}
table(spatial_coor$cell_type)
```

-----

```{r eval=FALSE}
ggplot(filter(spatial_coor,
              cell_type %in% c("B", "CD4")),
       aes(x = x, y = y, col = cell_type)) +
  geom_point() +
  facet_grid( . ~ cell_type) +
  guides(col = FALSE)
```


-----

```{r warning=FALSE, echo=FALSE}
ggplot(filter(spatial_coor,
              cell_type %in% c("B",  "CD4")),
       aes(x = x, y = y, col = cell_type)) +
  geom_point() +
  facet_grid( . ~ cell_type) +
  guides(col = FALSE)
```

-----

# Using the `spatstat` package

```{r}
ln <- with(spatial_coor,
ppp(x = x, y = y, marks = cell_type, 
    xrange = range(x), yrange = range(y)))
ln
```

-----

## Convex hull

```{r echo=FALSE, out.width="80%"}
coords <- cbind(ln$x, ln$y) 
chull <- convhulln(coords)

pidx = integer(nrow(chull) + 1) 
pidx[1:2] = chull[1, ]
chull[1, ] = NA

for(j in 3:length(pidx)) {
  wh = which(chull == pidx[j-1], arr.ind = TRUE)
  stopifnot(nrow(wh )== 1)
  wh[, "col"] = 3 - wh[, "col"] ## 2->1, 1->2 
  pidx[j] = chull[wh]
  chull[wh[, "row"], ] = NA
}
pidx = rev(pidx)

ggplot(tibble(x = ln$x, y = ln$y)[pidx, ], aes(x = x, y = y))+
  geom_point() + geom_path() + coord_fixed()

```

-----

## First order effects: intensity

```{r eval=FALSE}
d <- density(subset(ln, marks == "Macrophages"),
             edge=TRUE,
             diggle=TRUE)
plot(d)
```

-----

```{r echo=FALSE, out.width="90%"}
d <- density(subset(ln, marks == "Macrophages"),
             edge=TRUE,
             diggle=TRUE)
plot(d)
```


-----

# UMAP on protein expression

Dimension reduction technique -- could be used for visualisation

-----

Let’s load the umap package and apply the UMAP transformation

```{r include=FALSE}
library("umap")
library("M3C")
```


```{r}
assays.umap = umap(assay(subject_1))
```

-----

The output is an object and we can get a summary of its contents by printing it

```{r echo=FALSE}
assays.umap
```

-----


```{r message=FALSE, out.width="80%"}
umap(assay(subject_1),
     labels=colData(subject_1)$cell_type,
     legendtitle = "Cell Type")
```

-----

# References

[Easy quick PCA analysis in R](https://intobioinformatics.wordpress.com/2019/05/22/easy-quick-pca-analysis-in-r/)

[lmcinnes/umap](https://github.com/lmcinnes/umap)

[Modern Statistics for Modern Biology](http://web.stanford.edu/class/bios221/book/Chap-Images.html)

[Running UMAP for data visualisation in R](https://www.r-bloggers.com/2019/06/running-umap-for-data-visualisation-in-r/)

[UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction](https://arxiv.org/abs/1802.03426)

