---
title: "Week 6/7 - Incorporating Spatial Information with MIBI-TOF Dataset"
author: "Gheeda Mourtada"
date: "16/06/2021"
output: pdf_document
---

# Outline

- Points Drawn from _A Structured Tumor-Immune Microenvironment in Triple Negative Breast Cancer Revealed by Multiplexed Ion Beam Imaging_

- Creating `SpatialExperiment` for MIBI-TOF data

- Using `SpatialExperiment` object for creating heatmaps


## Introductory Notes

_MIBI-TOF:_ multiplexed ion beam imaging by time-of-flight 

**Goal:** understanding tumor-immune environment; learning more about immune phenotypes, compositions, and interactions with tumor cells.

**Results:** quantified _in-situ_ expression of 36 proteins in 41 patients.


\newpage

# SpatialExperiment for MIBI-TOF data

```{r include=FALSE}
library(dplyr)
library(SpatialExperiment)
```
```{r}
load("~/Downloads/mibiSCE.rda")
mibi.sce
```

```{r}
rowMeans(assay(mibi.sce))
```

\newpage


```{r}
rowSds(assay(mibi.sce))
```


The channels that were collected across cells in the experiment

```{r}
rownames(mibi.sce)
```

\newpage

Using the binary attribute is_protein for data filtering

```{r}
mibi.sce_proteins <- mibi.sce[rowData(mibi.sce)$is_protein == 1,]
mibi.sce_proteins
```

Removing the `mibi.sce` object from environment and using the `mibi.sce_proteins` moving forward

```{r}
rm(mibi.sce)
```


Identifying abundance of each cell type in `tumor_group`

```{r}
round(table(mibi.sce_proteins$tumor_group)/ncol(mibi.sce_proteins),2)
```


Identifying abundance of each cell type in `immune_group`

```{r}
round(table(mibi.sce_proteins$immune_group)/ncol(mibi.sce_proteins),2)
```

\newpage

Keratin +ve tumor cells and immune cells

```{r}
table(colData(mibi.sce_proteins)$tumor_group)
```


```{r}
table(colData(mibi.sce_proteins)$tumor_group, colData(mibi.sce_proteins)$immune_group)
```

\newpage

\footnotesize
```{r}
mibi.sce_proteins <- mibi.sce_proteins[,
                                       colData(mibi.sce_proteins)$tumor_group %in% c("Immune",
                                                                                       "Keratin-positive tumor")]
```
\normalsize

\newpage

```{r}
mibi <- mibi.sce_proteins
mibi
```


Dropping columns with no protein expression

```{r}
mibi <- mibi[, colSums(assay(mibi)) > 0]
mibi
```

\newpage

Dropping rows with no expression across columns

```{r}
mibi <- mibi[rowSums(assay(mibi)) > 0, ]
mibi
```


```{r}
summary(as.vector(assay(mibi)))
hist(as.vector(assay(mibi)), breaks = 100)
```

\newpage

Removing outlier expression

```{r}
rm_outliers <- apply(assay(mibi), 2, function(x){ sum(abs(x) > 5) >= 1 })
mibi <- mibi[, which(!rm_outliers)]
mibi
```

Adding cell IDs

```{r}
colData(mibi)$cell_id <- paste0("mibi_", seq(1, dim(mibi)[2]))
colnames(mibi) <- colData(mibi)$cell_id
colData(mibi)$DONOR_NO <- colData(mibi)$DONOR_NO %>% as.character()
```


\newpage

# Creating `SpatialExperiment` object from processed `SingleCellExperiment` object

```{r}
SE <- SpatialExperiment(
  rowData = rowData(mibi),
  colData = colData(mibi),
  assays = assay(mibi))
SE
```

\newpage

# Exploring: extraction of spatial information from tiff files

```{r include=FALSE}
library(plyr)
library(raster)
library(rgdal)
```

Saving the .tiff files as raster objects

```{r message=FALSE}
setwd("~/Downloads")
ras_obj<-raster('TNBC_shareCellData/p2_labeledcellData.tiff')
ras_obj
```

```{r}
plot(ras_obj, main = strsplit('TNBC_shareCellData/p2_labeledcellData.tiff','/')[[1]][2])
```

\newpage

```{r}
tiff_file_list<- list.files(path="TNBC_shareCellData/",
                            pattern = ".tiff")

id <- 2

str_name<- paste("TNBC_shareCellData/",tiff_file_list[id], sep = "")

sample_id <- as.numeric(gsub("p", "", gsub("_labeledcellData.tiff", "", tiff_file_list[id])))
sample_id
```

Creating a RasterLayer object from `str_name` 

```{r}
r <- raster(str_name)
r
```

\newpage

Matching the patients' _(i.e. sample)_ IDs from the tiff files to those in the `mibi` object

```{r}
mibi_sample_id_2 <- mibi[, mibi$SampleID == sample_id]
mibi_sample_id_2


df_r <- data.frame(rasterToPoints(r))
colnames(df_r) <- c("X", "Y", "values")
head(df_r)
```

\newpage

\footnotesize
```{r}
noise_not_in_mibi <- unique(df_r$values[!df_r$values %in% mibi_sample_id_2$cellLabelInImage])

# Adding cell_type column
cell_type <- ifelse(mibi_sample_id_2$immune_group != "not immune", mibi_sample_id_2$immune_group, "Tumor")
mibi_sample_id_2$cell_type <- cell_type

# Mapping values from those that are in noise_not_in_mibi to Background
cell_label_with_bg <- mapvalues((df_r$values), 
                                from = noise_not_in_mibi, 
                                to = rep("Background", length(noise_not_in_mibi)))

# Mapping values which are not Background to cell_type
cell_label_with_bg <- mapvalues((cell_label_with_bg), 
                                from = mibi_sample_id_2$cellLabelInImage, 
                                to = mibi_sample_id_2$cell_type)

df_r$cell_label_with_bg <- cell_label_with_bg

df_r$cell_label_with_bg <- factor(df_r$cell_label_with_bg)

df_r$cell_label_with_bg <- factor(df_r$cell_label_with_bg,
                                   levels = c(levels(df_r$cell_label_with_bg)[-which(levels(df_r$cell_label_with_bg) == "Background")], "Background"))
```
\normalsize

\newpage

```{r}
library(ggplot2)
library(ggthemes)

manual_col_sample_id <- tableau_color_pal("Classic 20")(length(unique(df_r$cell_label_with_bg))-1)
manual_col_sample_id <- c(manual_col_sample_id, "black")

ggplot(NULL) + 
  geom_raster(data = df_r, 
              aes(X, Y, fill = cell_label_with_bg)) +
  theme_minimal() +
  scale_fill_manual(values = manual_col_sample_id) +
  coord_quickmap() +
  theme(aspect.ratio = 1, legend.position = "right") +
  labs(fill = "Cell types")
```


\newpage

# Heatmaps from our previously created `SpatialExperiment` Object

## Plotting two categorical variables

```{r eval=FALSE}
col_SE<-data.frame(colData(SE))

anno_col = data.frame( TumorGroup=col_SE$tumor_group,
                       ImmuneGroup=col_SE$immune_group,
                       DonorNO=col_SE$DONOR_NO)

rownames(anno_col)<-colnames(assay(SE))

pheatmap::pheatmap(assay(SE),
                   annotation_col = anno_col,
                   cluster_rows = FALSE, 
                   cluster_cols = FALSE, 
                   fontsize_row = 8,
                   show_colnames = FALSE)
```

\newpage

```{r echo=FALSE}
col_SE<-data.frame(colData(SE))

anno_col = data.frame( TumorGroup=col_SE$tumor_group,
                       ImmuneGroup=col_SE$immune_group,
                       DonorNO=col_SE$DONOR_NO)

rownames(anno_col)<-colnames(assay(SE))

p<-pheatmap::pheatmap(assay(SE),
                   annotation_col = anno_col,
                   cluster_rows = FALSE, 
                   cluster_cols = FALSE, 
                   fontsize_row = 8,
                   show_colnames = FALSE)
p
ggsave("SE.png",p)
```

\newpage

## Spatial location of cells
```{r include=FALSE}
library(tidyverse)
```

\footnotesize
```{r}
setwd("~/Downloads")
tiff_file_list<- list.files(path="TNBC_shareCellData/",
                            pattern = ".tiff")[1:41]
tiff_file_ID <- as.list(seq_along(tiff_file_list))

SpatialCoordGetter <- function(id){
  
  # Similarly to the previously created "for" loop, we extract the sample IDs from the tiff files and create raster object from it
  sample_id <- as.numeric(gsub("p", "", gsub("_labeledcellData.tiff", "", tiff_file_list[id])))
  str_name <- paste("TNBC_shareCellData/",tiff_file_list[id],
                    sep = "")
  r <- raster(str_name)
  
  # creating dataframe which includes spatial information
  df_rP <- data.frame(rasterToPoints(r))
  
  # changing "value" column to "cellLabel"
  colnames(df_rP)[3] <- "cellLabel"
  
  # compute centroid x coordinate of each cell; first column of our dataframe (median of each cell using "cellLabel" as the list of grouping elements)
  centroid_x <- aggregate(df_rP[,1], 
    by = list(df_rP[,3]), FUN = median) %>%
     dplyr::rename('cellLabel' = Group.1)  
  # new name = old name; we get a data frame with cell label as the first column and x centroid as the second

  # compute centroid y coordinate of each cell; second column of our dataframe (similarly to calculating x centroid)
  centroid_y <- aggregate(df_rP[,2], 
    by = list(df_rP[,3]), FUN = median) %>% 
    dplyr::rename('y' = x, 'cellLabel' = Group.1) #renaming columns of dataframe

  # creating a data frame with cellLabel, x, and y, and adding a sample id column
  centroid_xy <- left_join(centroid_x, centroid_y, 
                           by = "cellLabel")
  
  centroid_xy <- centroid_xy %>% mutate(sample_id = rep(sample_id,
                                                        nrow(centroid_xy)))
  
  return(centroid_xy)
}

#binding the data frames of centorid of cell labels for all subjects
centroid_xy_df <- map_dfr(tiff_file_ID, SpatialCoordGetter)

# adding centroid_x and centroid_y to SE (spatial experiment object) while preserving the order of SampleID and cellLabelInImage
mibi_df <- data.frame(colData(SE))
mibi_df <- left_join(mibi_df,
                     centroid_xy_df,
                     by = c("SampleID" = "sample_id", "cellLabelInImage" = "cellLabel"))

SE$centroid_x <- mibi_df$x
SE$centroid_y <- mibi_df$y

```
\normalsize

\newpage

# Adding `spatialCoords` to our previously created `SpatialExperiment` object

```{r}
spatial_coords <- as.matrix(data.frame(
  x = SE$centroid_x, 
  y = SE$centroid_y
  ))

spatialCoords(SE) <- spatial_coords
SE
save(SE,file = "se.rds")
```

\newpage


```{r eval=FALSE}
FurtherProcessing<- function(id){
  
  str_name<- paste("TNBC_shareCellData/",
                   tiff_file_list[id], sep = "")
  sample_id <- as.numeric(gsub("p", "",
                               gsub("_labeledcellData.tiff", "",
                                    tiff_file_list[id])))
   r <- raster(str_name)
   
   df<-data.frame(rasterToPoints(r))
   colnames(df) <- c("X", "Y", "values")
      
   # matching sample IDs from tiff files to those in the SCE object mibi
  mibi_sample_id <- mibi[, mibi$SampleID == sample_id]

  # cell labels not in SCE
  noise_not_in_mibi <- unique(df$values[!df$values %in% mibi_sample_id$cellLabelInImage])

  # Adding cell_type column
  cell_type <- ifelse(mibi_sample_id$immune_group != "not immune", mibi_sample_id$immune_group, "Tumor")
  mibi_sample_id$cell_type <- cell_type

  # Mapping values from those that are in noise_not_in_mibi to Background
  cell_label_with_bg <- mapvalues((df$values),
                                  from = noise_not_in_mibi,
                                  to = rep("Background",
                                           length(noise_not_in_mibi)))

    # Mapping values which are not Background to cell_type
  cell_label_with_bg <- mapvalues((cell_label_with_bg),
                                  from = mibi_sample_id$cellLabelInImage,
                                  to = mibi_sample_id$cell_type)

  df$cell_label_with_bg <- cell_label_with_bg
  df$cell_label_with_bg <- factor(df$cell_label_with_bg)
  df$cell_label_with_bg <- factor(df$cell_label_with_bg,
                                   levels = c(levels(df$cell_label_with_bg)[-which(levels(df$cell_label_with_bg) == "Background")], "Background"))
  
  return(df)
  
}

# binding the data frames including cell types for all subjects
df_41 <- map_dfr(tiff_file_ID, FurtherProcessing)

centroid_xy_df$cell_type<-centroid_xy_df$x

# mapping the cell type according to    
  centroid_xy_df$cell_type<-mapvalues(centroid_xy_df$cell_type,
                                      from = df_41$X,
                                      to = df_41$cell_label_with_bg)

```


\newpage

<!-- Attempt to: create `mibi_sample_id_` for all subjects separately and create the data frame for all subjects from `mibi_sample_id` -->

<!-- \footnotesize -->
<!-- ```{r eval=FALSE} -->
<!-- manual_col_sample_id <- c("B"="#1f77b4", -->
<!--                           "CD3"="#aec7e8", -->
<!--                           "CD4"="#ff7f0e", -->
<!--                           "CD8"="#ffbb78", -->
<!--                           "DC"="#2ca02c", -->
<!--                           "DC/Mono"="#98df8a", -->
<!--                           "Macrophages"="#c49c94", -->
<!--                           "Mono/Neu"="#ff9896", -->
<!--                           "Neutrophils"="#9467bd", -->
<!--                           "NK"="#c5b0d5", -->
<!--                           "Other immune"="#8c564b", -->
<!--                           "Tregs"="#e377c2", -->
<!--                           "Tumor"="#d62728", -->
<!--                           "Background"="black") -->

<!-- tiff_file_list<- list.files(path="TNBC_shareCellData/", -->
<!--                             pattern = ".tiff") -->
<!-- id<-c(1:41) -->

<!-- for (val in id){ -->
<!-- str_name_id<- paste("TNBC_shareCellData/",tiff_file_list[val], sep = "") -->

<!-- sample_id_each <- as.numeric(gsub("p", "", gsub("_labeledcellData.tiff", "", tiff_file_list[val]))) -->
<!-- r <- raster(str_name_id) -->
<!-- mibi_sample_id <- mibi[, mibi$SampleID == sample_id_each] -->

<!-- saveRDS(mibi_sample_id, paste(sample_id_each,".rds")) -->

<!-- # Reading one of the saved SingleCellExperiment objects -->

<!-- sce_file_list<- list.files(pattern = ".rds") -->

<!-- df_rP <- data.frame(rasterToPoints(r)) -->
<!-- colnames(df_rP) <- c("X", "Y", "values") -->
<!-- head(df_rP) -->

<!-- noise_not_in_mibi <- unique(df_rP$values[!df_rP$values %in% mibi_sample_id$cellLabelInImage]) -->

<!-- # Adding cell_type column -->
<!-- cell_type <- ifelse(mibi_sample_id$immune_group != "not immune", mibi_sample_id$immune_group, "Tumor") -->
<!-- mibi_sample_id$cell_type <- cell_type -->

<!-- # Mapping values from those that are in noise_not_in_mibi to Background -->
<!-- cell_label_with_bg <- mapvalues((df_rP$values),  -->
<!--                                 from = noise_not_in_mibi,  -->
<!--                                 to = rep("Background", length(noise_not_in_mibi))) -->

<!-- # Mapping values which are not Background to cell_type -->
<!-- cell_label_with_bg <- mapvalues((cell_label_with_bg),  -->
<!--                                 from = mibi_sample_id$cellLabelInImage,  -->
<!--                                 to = mibi_sample_id$cell_type) -->

<!-- df_rP$cell_label_with_bg <- cell_label_with_bg -->

<!-- df_rP$cell_label_with_bg <- factor(df_rP$cell_label_with_bg) -->

<!-- df_rP$cell_label_with_bg <- factor(df_rP$cell_label_with_bg, -->
<!--                                    levels = c(levels(df_rP$cell_label_with_bg)[-which(levels(df_rP$cell_label_with_bg) == "Background")], "Background")) -->

<!-- df_rP$subjectID <- sample_id_each  -->

<!-- assign( paste("df_rP_sample_ID", val, sep = "_") , df_rP)   -->

<!-- p<-ggplot(NULL) +  -->
<!--   geom_raster(data = df_rP,  -->
<!--               aes(X, Y, fill = cell_label_with_bg)) + -->
<!--   theme_minimal() + -->
<!--   scale_fill_manual(values = manual_col_sample_id) + -->
<!--   coord_quickmap() + -->
<!--   theme(aspect.ratio = 1, legend.position = "right") + -->
<!--   labs(fill = "Cell types") -->

<!-- ggsave(paste0("df_rP_sample_ID_plot", val, ".png"), p) -->

<!-- } -->

<!-- readRDS("~/Downloads/1 .rds") -->
<!-- ``` -->
<!-- \normalsize -->

\newpage

# References

[A Structured Tumor-Immune Microenvironment in Triple Negative Breast Cancer Revealed by Multiplexed Ion Beam Imaging](https://www.cell.com/action/showPdf?pii=S0092-8674%2818%2931100-0)

[Building SpatialExperiment objects](https://bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html)

Creating SpatialExperiment from SingleCellExperiment and images _Dr. Jeganathan_

[Efficiently bind multiple data frames by row and column](https://dplyr.tidyverse.org/reference/bind.html)

[Raster data manipulation](https://rspatial.org/raster/spatial/8-rastermanip.html)

[Raster usage for accessing coordinate labels](https://github.com/BIRSBiointegration/Hackathon/tree/master/sc-targeted-proteomics/spatial_data)

Integrating proteomics data _Dr. Jeganathan_


[MIBI Datasets](https://www.angelolab.com/mibi-data)

[Pheatmap Draws Pretty Heatmaps](https://towardsdatascience.com/pheatmap-draws-pretty-heatmaps-483dab9a3cc)

[SpatialExperiment-methods: Methods for spatial attributes](https://rdrr.io/github/drighelli/VisiumExperiment/man/SpatialExperiment-methods.html)
